Structured Versus Decorated Cospans

### 2021-06-15 cospans_new.tex
changes in and around Thms. 2.1 and 2.2

We start with decorated cospans, proving Theorems \ref{thm:decorated_cospans_1} and \ref{thm:decorated_cospans_2} by applying Shulman's result to a particular monoidal bifibration built
from a lax monoidal pseudofunctor $F \maps (A,+) \to (\Cat,\times)$.    The Grothendieck construction gives a bijection between pseudofunctors $F \maps \A \to \Cat$ and opfibrations $U \maps \inta F \to \A$.  We need two refinements of this construction: one that gives monoidal obfibrations, and one that gives symmetric monoidal opfibrations.  We explain the Grothendieck construction and all these related concepts in \cref{subsec:fibrations}; here we simply recall the result we need.

@@

\begin{proof}[Proof of \cref{thm:decorated_cospans_1}] 
To obtain the desired double category we apply \cref{lem:shulman} to a bifibration $\Phi \maps \A/U \to \A$ which we now define.   Let $U \maps \inta F \to \A$ be the monoidal opfibration associated to the lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat, \times)$.  As in \cref{subsec:fibrations},
\begin{itemize}
\item an object of $\inta F$ is a pair $(b,p)$ where $b \in \A$ and $p \in F(b)$;
\item a morphism in $\inta F$ from $(b,p)$ to $(b',p')$ is a pair $(f,\zeta)$
where $g \maps b \to b'$ is a morphism in $\A$ and $\zeta \maps F(g)(p) \to p'$ is
a morphism in $F(b')$.
\end{itemize}
The functor $U$ maps objects and morphisms to their first component, so in the comma
category $\A/U$, 
\begin{itemize}
\item an object is a triple $(a,x,a \xrightarrow{i} U(x))$ with $a \in \A$ and $x=(b,p) \in \inta F$;
\item a morphism from $(a,x,a \xrightarrow{i} U(x))$ to $(a',x',a' \xrightarrow{i'} U(x'))$ is a pair $(f,\theta)$ where $f \maps a \to a'$ is a morphism in $\A$ and $\theta \maps x \to x'$ is a morphism in $\inta F$ such that the following square commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x')$.};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge node[right]{$i'$} (D)
(C) edge node[below]{$U(\theta)$} (D);
\end{tikzpicture}
\]
\end{itemize}
The functor $\Phi \maps \A/U \to \A$ is defined on objects by $\Phi(a,x,a \xrightarrow{i} U(x))=a$ and on morphisms by $\Phi(f,\theta)=f$. We have a monoidal structure on $\A/U$ where the tensor product of two objects $(a_1,x_1,a_1 \xrightarrow{i_1} U(x_1))$ and $(a_2,x_2,a_2 \xrightarrow{i_2} U(x_2))$ is given by
\[  (a_1,x_1,a_1 \xrightarrow{i_1} U(x_1)) \otimes (a_2,x_2,a_2 \xrightarrow{i_2} U(x_2)) = (a_1+a_2,x_1 \otimes x_2,a_1+a_2 \xrightarrow{i_1+i_2} U(x_1 \otimes x_2)) \]
where we have left the equality $U(x_1)+U(x_2) = U(x_1 \otimes x_2)$ implicit in the above composite, for reasons of space. The monoidal unit is given by $(0_\A,0_\X,0_\A \to U(0_\X))$. It is easy to see that the functor $\Phi$ is a strict monoidal using this tensor product.

One can easily check that $\Phi$ is a fibration.  Given a morphism $f \maps a \to b$ in $\A$ and an object $(b,y,b \xrightarrow{i} U(y)) \in \A/U$ over $b$, a cartesian lifting of $f$ is given by 
\[ (f,\id) \maps (a,y,a \xrightarrow{i \circ f} U(y)) \, \to \, (b,y,b \xrightarrow{i} U(y)). \]
Denoting this cartesian lifting by $\textnormal{cart}(f)$, the tensor product of $\A/U$ strictly preserves such cartesian liftings in the sense that $\textnormal{cart}(f_1) \otimes \textnormal{cart}(f_2)=\textnormal{cart}(f_1+f_2)$.

One can also check that the functor $\Phi$ is an opfibration, but this takes a bit more work. Given a morphism $f \maps a \to a'$ in $\A$ and an object $(a,x,a \xrightarrow{i} U(x)) \in \A/U$ over $a$, we shall show that a cocartesian lifting of $f$ is given by 
\[ (f,\theta) \maps (a,x,a \xrightarrow{i} U(x)) \, \to \, (a',x', a' \xrightarrow{i'} U(x')). \]
%%% Kenny: Just an observation: we're using a morphism f:a-->b for the fibration part and a morphism f:a-->a' for the opfibration part.
Here $i'$ arises from the following pushout square in $\A$:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x)+_a a'$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge [dashed] node[right]{$i'$} (D)
(C) edge [dashed] node[below]{$g$} (D);
\end{tikzpicture}
\]
The object $x' \in \inta F$ is defined as follows.  Recall that $x \in \inta F$ is a
pair $(b,p)$ with $b\in \A$ and $p \in F(b)$, and note that $U(x) = b$.   Define 
$x'$ to be the pair $(b',p')$ where $b' = U(x)+_a a'$ is the pushout in the above
diagram and $p' = F(g)(p)$.  Then $U(x') = b'$ so we have 
$i' \maps a' \to U(x')$ as desired.  Finally, we define $\theta \maps x \to x'$ to be 
the morphism in $\inta F$ given by the pair $(g,\id)$ where $g \maps b \to b'$ is
given in the above diagram and $\id \maps F(g)(p) \to p'$ is the identity.

{\john To show that $(f,\theta)$ is a cocartesian lifting of $f$, suppose...
Then use this picture somehow:}
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (2.5,0) {$c$};
\node (C) at (1.5,1) {$a'$};
\node (D) at (0,-2) {$U(x)$};
\node (E) at (2.5,-2) {$U(y)$};
\node (F) at (1.5,-1) {$U(x') = U(x)+_a a'$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$\theta$} (F)
(F) edge [dashed] node [right] {$\exists ! \psi$} (E)
(D) edge node [above] {$U(g')$} (E)
(A) edge node [left] {$i$} (D)
(C) edge node [right,near end] {$i'$} (F)
(B) edge node {$$} (E)
(A) edge node[above]{$f'$} (B)
(C) edge node [right] {$p$} (B)
(A) edge node[left]{$f$} (C);
%(B) edge [dashed] node[right]{$$} (D)
%(C) edge [dashed] node[above]{$$} (D);
\end{tikzpicture}
\]

@@

\end{tikzpicture}
\]
and consequently a natural isomorphism between our two objects of $(\A/U)_c$, showing that $\Phi$ satisfies the Beck--Chevalley condition.
\end{proof}

\begin{proof}[Proof of \cref{thm:decorated_cospans_2}] 
When the lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat, \times)$ of \cref{thm:decorated_cospans_1} is symmetric, the monoidal bifibration $\Phi \maps \A/U \to A$
is symmetric, so \cref{lem:shulman} implies that the double category $F\Csp$ becomes symmetric
monoidal as described.
\end{proof}

>>>>

We start with decorated cospans, proving Theorems \ref{thm:decorated_cospans_1} and \ref{thm:decorated_cospans_2} by applying Shulman's result to a particular monoidal bifibration built
from a lax monoidal pseudofunctor $F \maps (A,+) \to (\Cat,\times)$.    The Grothendieck construction gives a bijection between pseudofunctors $F \maps \A \to \Cat$ and opfibrations $U \maps \inta F \to \A$.  We need two refinements of this construction: one that gives monoidal ofibrations, and one that gives symmetric monoidal opfibrations.  We explain the Grothendieck construction and all these related concepts in \cref{subsec:fibrations}; here we simply recall the result we need.

@@

\begin{proof}[Proof of \cref{thm:decorated_cospans_1}] 
To obtain the desired double category we apply \cref{lem:shulman} to a bifibration $\Phi \maps \A/U \to \A$ which we now define.   Let $U \maps \inta F \to \A$ be the monoidal opfibration associated to the lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat, \times)$.  As in \cref{def:GrothCat},
\begin{itemize}
\item an object of $\inta F$ is a pair $(b,p)$ where $b \in \A$ and $p \in F(b)$;
\item a morphism in $\inta F$ from $(b,p)$ to $(b',p')$ is a pair $(f,\zeta)$
where $g \maps b \to b'$ is a morphism in $\A$ and $\zeta \maps F(g)(p) \to p'$ is
a morphism in $F(b')$.
\end{itemize}
The functor $U$ maps objects and morphisms to their first component, so in the comma
category $\A/U$, 
\begin{itemize}
\item an object is a triple $(a,x,a \xrightarrow{i} U(x))$ with $a \in \A$ and $x=(b,p) \in \inta F$;
\item a morphism from $(a,x,a \xrightarrow{i} U(x))$ to $(a',x',a' \xrightarrow{i'} U(x'))$ is a pair $(f,\theta)$ where $f \maps a \to a'$ is a morphism in $\A$ and $\theta \maps x \to x'$ is a morphism in $\inta F$ such that this square commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x')$.};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge node[right]{$i'$} (D)
(C) edge node[below]{$U(\theta)$} (D);
\end{tikzpicture}
\]
\end{itemize}
We define the functor $\Phi \maps \A/U \to \A$ on objects by $\Phi(a,x,a \xrightarrow{i} U(x))=a$ and on morphisms by $\Phi(f,\theta)=f$.

There is a monoidal structure on $\A/U$ where the tensor product of two objects $(a_1,x_1,a_1 \xrightarrow{i_1} U(x_1))$ and $(a_2,x_2,a_2 \xrightarrow{i_2} U(x_2))$ is given by
\[  (a_1,x_1,a_1 \xrightarrow{i_1} U(x_1)) \otimes (a_2,x_2,a_2 \xrightarrow{i_2} U(x_2)) = (a_1+a_2,x_1 \otimes x_2,a_1+a_2 \xrightarrow{i_1+i_2} U(x_1 \otimes x_2)) \]
where we have left the equality $U(x_1)+U(x_2) = U(x_1 \otimes x_2)$ implicit in the above composite, for reasons of space. The monoidal unit is given by $(0_\A,0_\X,0_\A \to U(0_\X))$. It is easy to see that the functor $\Phi$ is a strict monoidal using this tensor product.

One can check that $\Phi$ is a fibration.  Given a morphism $f \maps a \to a'$ in $\A$ and an object $(a',x',a' \xrightarrow{i'} U(x')) \in \A/U$ over $a'$, a cartesian lifting of $f$ is given by 
\[ (f,\id) \maps (a,x',a \xrightarrow{i' \circ f} U(x')) \, \to \, (a',x',a' \xrightarrow{i'} U(x')). \]
Denoting this cartesian lifting by $\textnormal{cart}(f)$, the tensor product of $\A/U$ strictly preserves such cartesian liftings in the sense that $\textnormal{cart}(f_1) \otimes \textnormal{cart}(f_2)=\textnormal{cart}(f_1+f_2)$.

One can also check that the functor $\Phi$ is an opfibration, but this takes a bit more work. Given a morphism $f \maps a \to a'$ in $\A$ and an object $(a,x,a \xrightarrow{i} U(x)) \in \A/U$ over $a$, we shall show that a cocartesian lifting of $f$ is given by 
\[ (f,\theta) \maps (a,x,a \xrightarrow{i} U(x)) \, \to \, (a',x', a' \xrightarrow{i'} U(x')). \]
where $i'$ arises from the following pushout square in $\A$:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1.5,0) {$a'$};
\node (C) at (0,-1) {$U(x)$};
\node (D) at (1.5,-1) {$U(x)+_a a'$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$f$} (B)
(A) edge node[left]{$i$} (C)
(B) edge [dashed] node[right]{$i'$} (D)
(C) edge [dashed] node[below]{$g$} (D);
\end{tikzpicture}
\]
The object $x' \in \inta F$ is defined as follows.  Recall that $x \in \inta F$ is a
pair $(b,p)$ with $b\in \A$ and $p \in F(b)$, and note that $U(x) = b$.   Define 
$x'$ to be the pair $(b',p')$ where $b' = U(x)+_a a'$ is the pushout in the above
diagram and $p' = F(g)(p)$.  Then $U(x') = b'$ so we have 
$i' \maps a' \to U(x')$ as desired.  Finally, define $\theta \maps x \to x'$ to be 
the morphism in $\inta F$ given by the pair $(g,\id)$ where $g \maps b \to b'$ is
given as above and $\id \maps F(g)(p) \to p'$ is the identity.

To show that $(f,\theta)$ is an cocartesian lifting of $f$, suppose

Then use this picture somehow:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (2.5,0) {$c$};
\node (C) at (1.5,1) {$a'$};
\node (D) at (0,-2) {$U(x)$};
\node (E) at (2.5,-2) {$U(y)$};
\node (F) at (1.5,-1) {$U(x') = U(x)+_a a'$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$\theta$} (F)
(F) edge [dashed] node [right] {$\exists ! \psi$} (E)
(D) edge node [above] {$U(g')$} (E)
(A) edge node [left] {$i$} (D)
(C) edge node [right,near end] {$i'$} (F)
(B) edge node {$$} (E)
(A) edge node[above]{$f'$} (B)
(C) edge node [right] {$p$} (B)
(A) edge node[left]{$f$} (C);
%(B) edge [dashed] node[right]{$$} (D)
%(C) edge [dashed] node[above]{$$} (D);
\end{tikzpicture}
\]

@@

\end{tikzpicture}
\]
and consequently a natural isomorphism between our two objects of $(\A/U)_c$, showing that $\Phi$ satisfies the Beck--Chevalley condition.   We then define $F\lCsp$ to be $\Fr(\Phi)$.
\end{proof}

\begin{proof}[Proof of \cref{thm:decorated_cospans_2}] 
When the lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat, \times)$ of \cref{thm:decorated_cospans_1} is symmetric, the monoidal bifibration $\Phi \maps \A/U \to A$
is symmetric, so \cref{lem:shulman} implies that the double category $F\lCsp = \Fr(\Phi)$ is symmetric monoidal as described.
\end{proof}
